---
title: "Constructing a factor from scratch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Constructing a factor from scratch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

Between a raw download of stock returns and a published asset-pricing result
sits a long chain of decisions: how to lag accounting data so a portfolio
formed today uses only information available today, where to draw the
breakpoints that separate "high" from "low" firms, whether to weight stocks
equally or by size, and when to rebalance. These choices rarely appear in a
paper's tables, yet they move the numbers. The goal of `tidyfinance` is to make
each choice an explicit, documented argument rather than a buried assumption.

This vignette walks through the whole chain once, end to end. We build a single
factor --- the *asset growth* factor of Cooper, Gulen, and Schill (2008), which
sorts firms on the year-over-year change in their total assets --- starting from
firm-level accounting and return panels and finishing with a long-short return
series and a cross-sectional asset-pricing test. The vignette has six steps: we
download the data, construct and lag the signal, align it with monthly returns,
sort stocks into portfolios, customize that sort, and finallsy run a Fama-MacBeth
regression.

## Following along without a WRDS subscription

The real asset growth factor is built from CRSP (stock returns) and Compustat
(accounting statements), both distributed through Wharton Research Data Services
(WRDS) and both requiring an institutional subscription. So that this vignette
builds anywhere --- on CRAN, on a continuous-integration runner, on a laptop
with no credentials and no network --- every download below uses
`domain = "Pseudo Data"` instead of `domain = "WRDS"`.

Pseudo data mirror the *schema* of the real WRDS panels exactly: the same column
names, types, and panel structure, and a
calibrated mix of exchanges and industries. Every line of cleaning, lagging, and
sorting code therefore runs unchanged. What pseudo data deliberately do **not**
carry is economic content. Returns are independent random draws, and the asset
growth signal is unrelated to subsequent returns by construction. **Every number
in this vignette is a simulation artifact** --- not just the cross-sectional
premia and slopes, but the levels too (the uniformly positive "excess" returns
reflect a built-in drift in the generator, not a market risk premium). Treat the
code as the lesson and the numbers as placeholders; to reproduce the published
premium, swap `"Pseudo Data"` for `"WRDS"` (after a one-time
`set_wrds_credentials()`) and rerun.

```{r}
#| label: load-packages
library(tidyfinance)
library(dplyr)
library(tidyr)
library(lubridate)
```

Two arguments control the simulated panels: `seed` fixes the random draws so the
document is reproducible, and `n_assets` sets the size of the
cross-section. We keep a
moderate universe here and a sample window that mirrors the empirical study.

```{r}
#| label: parameters
seed <- 1234
n_assets <- 500
date_start <- as.Date("1972-01-01")
date_end <- as.Date("2024-12-31")
```

## Step 1: Download the data

Constructing the factor needs two panels: monthly stock returns from CRSP and
annual accounting data from Compustat. `download_data()` is the single entry
point for both; the `domain` and `dataset` arguments pick the source.

Setting `add_ccm_links = TRUE` appends the CRSP-Compustat linking identifier
(`gvkey`) to the return panel, which we need to merge in accounting data later.

```{r}
#| label: download-crsp
crsp_monthly <- download_data(
  domain = "Pseudo Data",
  dataset = "crsp_monthly",
  start_date = date_start,
  end_date = date_end,
  add_ccm_links = TRUE,
  n_assets = n_assets,
  seed = seed
)
crsp_monthly
```

Each row is one stock-month, already cleaned: market capitalization (`mktcap`) and
its one-month lag (`mktcap_lag`) for value weighting, a readable `exchange` label,
the SIC industry code (`siccd`), and excess returns (`ret_excess`) net of the
risk-free rate. On real WRDS data `download_data()` also restricts the universe to
common equity, folds in delisting returns, and maps exchange and industry codes.
Here those columns are simulated, but the exchange and industry mix is calibrated
to real CRSP frequencies and `siccd` is drawn to match each firm's industry, so
the exchange and industry filters we apply in Step 5 still drop the intended firms.

Next we download annual Compustat statements. The `additional_columns` argument
pulls in any raw fields beyond the defaults; we ask for total assets (`at`) ---
the input to the asset growth signal --- and income before extraordinary items
(`ib`), which we use later as an earnings filter.

```{r}
#| label: download-compustat
compustat_annual <- download_data(
  domain = "Pseudo Data",
  dataset = "compustat_annual",
  start_date = date_start,
  end_date = date_end,
  additional_columns = c("at", "ib"),
  only_usd = TRUE,
  n_assets = n_assets,
  seed = seed
) |>
  select(gvkey, date, at, ib)
compustat_annual
```

## Step 2: Construct and lag the signal

Asset growth for firm $i$ in year $t$ is the year-over-year percentage change in
total assets,

$$
AG_{t,i} = \frac{AT_{t,i} - AT_{t-1,i}}{AT_{t-1,i}}.
$$

Computing it means lining up each firm's total assets with its own value one year
earlier. `add_lagged_columns()` does exactly that: grouped by firm (`gvkey`), it
attaches each firm's total assets from exactly one year before (`lag = years(1)`),
matching the lagged date directly. The same function also takes a `max_lag` to
instead search a window for the most recent value, which we use in the next step.
We then take the percentage change and set non-finite results (from a zero or
missing prior value) to `NA`.

```{r}
#| label: compute-signal
sorting_variable_data <- compustat_annual |>
  add_lagged_columns(
    cols = "at",
    lag = years(1),
    by = "gvkey"
  ) |>
  mutate(
    asset_growth = (at - at_lag) / at_lag,
    asset_growth = if_else(is.finite(asset_growth), asset_growth, NA_real_)
  ) |>
  select(gvkey, date, asset_growth, ib)
```

Before using the signal, it pays to look at it. `create_summary_statistics()`
returns the standard cross-sectional summary for one or more variables. We
winsorize first --- `winsorize()` pulls the most extreme 1% on each tail back to
the 1st and 99th percentiles --- so that a handful of firms with explosive
balance-sheet changes do not distort the picture.

```{r}
#| label: signal-summary
sorting_variable_data |>
  drop_na(asset_growth) |>
  mutate(asset_growth = winsorize(asset_growth, cut = 0.01)) |>
  create_summary_statistics(asset_growth, detail = TRUE) |>
  select(variable, n, mean, sd, q05, q50, q95) |>
  knitr::kable(
    digits = 3,
    caption = "Cross-sectional distribution of asset growth."
  )
```

Even after winsorizing, the mean sits above the median: asset growth is
right-skewed, with a tail of fast-expanding firms pulling the average up. That
skew is a property of the constructed variable and so survives in pseudo data ---
unlike any relationship between the signal and future returns, which does not.

## Step 3: Align the signal with returns

The signal is annual; returns are monthly. To form portfolios we need each
stock-month paired with the *most recently observable* value of asset growth ---
using only what an investor could have known at the formation date. This is where
look-ahead bias (slipping in information that was not yet public) creeps in if you are
not careful: the accounting for fiscal year $t$ is not public the instant the
year ends.

`join_lagged_values()` merges the annual signal onto the monthly panel through a
configurable lag window. Setting `ff_adjustment = TRUE` applies the Fama-French
(1993) timing convention: accounting data are assumed public roughly six months
after the fiscal year-end and held for the following twelve months. Because the
pseudo statements all carry a December fiscal year-end, the 7-to-18-month window
places each signal in the July-to-June holding period that the convention
prescribes. The window is anchored on the statement date, so for real firms with
other fiscal year-ends the exact calendar months shift accordingly.

```{r}
#| label: align-signal
sorting_data <- crsp_monthly |>
  join_lagged_values(
    new_data = sorting_variable_data,
    id_keys = "gvkey",
    min_lag = months(7),
    max_lag = months(18),
    ff_adjustment = TRUE
  ) |>
  select(
    date,
    permno,
    exchange,
    siccd,
    ret_excess,
    mktcap_lag,
    asset_growth,
    ib
  ) |>
  filter(date >= date_start + months(12) + months(18))
```

The final `filter()` drops the start of the sample, where the one-year growth
calculation and the lag window leave the signal mechanically missing. What
remains is a clean stock-month panel carrying excess returns, a value weight, and
a properly lagged signal.

## Step 4: Sort stocks into portfolios

A portfolio sort ranks stocks each period by the signal and splits them into
ordered groups; the spread between the extreme groups is the factor return.
`implement_portfolio_sort()` runs the whole sort in one call. It takes the panel,
the sorting variable, the sorting method, and a `portfolio_sort_options` object
that bundles how breakpoints are formed.

For the baseline we sort into quintiles, computing breakpoints from NYSE-listed
stocks only --- a standard convention that keeps the many small NASDAQ firms from
dominating the cutoffs --- and rebalance every July to match the Fama-French
timing we used when lagging the signal.

```{r}
#| label: baseline-sort
options_baseline <- portfolio_sort_options(
  breakpoint_options_main = breakpoint_options(
    n_portfolios = 5,
    breakpoints_exchanges = "NYSE"
  )
)

portfolio_returns <- implement_portfolio_sort(
  data = sorting_data,
  sorting_variables = "asset_growth",
  sorting_method = "univariate",
  rebalancing_month = 7,
  portfolio_sort_options = options_baseline
)
portfolio_returns
```

This baseline call named only two things; everything else came from a documented
default. That is the point: a default call reproduces standard practice, and every
construction choice that papers usually leave implicit --- an 80th-percentile cap
for the capped value-weighted return, no minimum portfolio size, the standard
CRSP/Compustat column names --- is here an argument you can see and change. The
output reports three weighting schemes per portfolio-month: value-weighted
(`ret_excess_vw`), equal-weighted (`ret_excess_ew`), and capped value-weighted
(`ret_excess_vw_capped`).

The factor itself is the long-short spread. `compute_long_short_returns()` takes
the difference between the extreme portfolios; `direction = "bottom_minus_top"`
goes long low-asset-growth firms and short high-asset-growth firms, the sign of
the documented premium.

```{r}
#| label: long-short
factor_returns <- portfolio_returns |>
  compute_long_short_returns(direction = "bottom_minus_top")
```

```{r}
#| label: portfolio-table
portfolio_summary <- bind_rows(
  portfolio_returns |>
    mutate(portfolio = as.character(portfolio)),
  factor_returns |>
    mutate(portfolio = "Long-short")
) |>
  group_by(portfolio) |>
  summarize(
    mean_vw = mean(ret_excess_vw, na.rm = TRUE) * 12,
    mean_ew = mean(ret_excess_ew, na.rm = TRUE) * 12,
    .groups = "drop"
  )

portfolio_summary |>
  knitr::kable(
    digits = 4,
    col.names = c("Portfolio", "Value-weighted", "Equal-weighted"),
    caption = paste(
      "Annualized mean excess returns by asset growth quintile, plus the",
      "long-short spread. On real data the value-weighted mean declines",
      "across quintiles and the long-short row is a sizable premium; here",
      "the quintile ordering and the long-short value (and its sign) are",
      "sampling noise."
    )
  )
```

On real WRDS data the value-weighted means decline monotonically from the low- to
the high-asset-growth quintile, and the long-short spread reproduces the negative
premium of Cooper, Gulen, and Schill (2008). On the pseudo panel there is no
systematic decline: the differences across quintiles, any apparent ordering, and
the long-short spread (whatever its sign on a given run) are all sampling noise, as
they must be.

## Step 5: Customize the sort

The baseline took nearly every default. A realistic specification overrides
several at once, and `portfolio_sort_options` keeps them all in one place. Here we
exclude financial and utility firms and firms with non-positive earnings, ignore
the smallest 20% of NYSE firms when setting breakpoints, and add a second sort on
size: a *dependent bivariate* sort that controls for size by first bucketing on
it, then sorting on asset growth within each bucket.

```{r}
#| label: customized-sort
options_customized <- portfolio_sort_options(
  filter_options = filter_options(
    exclude_financials = TRUE,
    exclude_utilities = TRUE,
    exclude_negative_earnings = TRUE
  ),
  breakpoint_options_main = breakpoint_options(
    n_portfolios = 5,
    breakpoints_exchanges = "NYSE",
    breakpoints_min_size_threshold = 0.2
  ),
  breakpoint_options_secondary = breakpoint_options(
    n_portfolios = 2,
    breakpoints_exchanges = "NYSE"
  )
)

factor_returns_customized <- implement_portfolio_sort(
  data = sorting_data,
  sorting_variables = c("asset_growth", "mktcap_lag"),
  sorting_method = "bivariate-dependent",
  rebalancing_month = 7,
  portfolio_sort_options = options_customized,
  min_portfolio_size = 10,
  quiet = TRUE
) |>
  compute_long_short_returns(direction = "bottom_minus_top")
```

Moving from a univariate to a size-controlled bivariate sort was a one-word change
to `sorting_method`; the surrounding pipeline is identical. That is the point of
exposing construction choices as arguments. The lower-level building blocks
(`compute_breakpoints()`, `assign_portfolio()`, `compute_portfolio_returns()`)
remain available when you need finer control over one stage.

## Step 6: Run an asset-pricing test

A long-short return is one verdict on whether a signal predicts the cross-section
of returns. A Fama-MacBeth regression (Fama and MacBeth, 1973) gives a complementary one: it runs a
cross-sectional regression of returns on the signal period by period, then
averages the slopes into a risk premium with Newey-West standard errors (Newey and West, 1987).
`estimate_fama_macbeth()` does this from a panel and a model formula.

We take the winsoized signal within each month and add log market capitalization as a
size control.

```{r}
#| label: fama-macbeth
fm_data <- sorting_data |>
  drop_na(asset_growth, mktcap_lag, ret_excess) |>
  group_by(date) |>
  mutate(
    asset_growth = winsorize(asset_growth, cut = 0.01),
    log_mktcap = log(mktcap_lag)
  ) |>
  ungroup()

fm_results <- estimate_fama_macbeth(
  data = fm_data,
  model = "ret_excess ~ asset_growth + log_mktcap"
)

fm_results |>
  knitr::kable(
    digits = 4,
    caption = paste(
      "Fama-MacBeth regression of monthly excess returns on lagged asset",
      "growth and log market cap. On real data the asset growth slope is",
      "negative and significant; on pseudo data it is indistinguishable",
      "from zero. The large, 'significant' intercept is a mechanical",
      "artifact of the simulated returns' built-in positive drift, not an",
      "alpha."
    )
  )
```

On real data the slope on `asset_growth` is negative and significant, matching the
portfolio-sort evidence: high-asset-growth firms earn lower subsequent returns. On
the pseudo panel the slope is statistical noise. The intercept row carries a large
$t$-statistic here, but that reflects only the generator's built-in positive return
drift, not a real average return. Switching the standard-error
treatment (`vcov = "iid"`) or adding a regressor is a one-argument change.

## Where to go next

You have now run the complete `tidyfinance` workflow: download, construct and lag
the signal, align it with returns, sort, customize, and test. Every stage exposed
its decisions as arguments with sensible defaults, so the baseline reproduces
standard practice and any variation is a small edit.

To reproduce the *published* asset growth premium, replace `"Pseudo Data"` with
`"WRDS"` in the two `download_data()` calls and run `set_wrds_credentials()` once
beforehand; nothing else in the pipeline changes. For the underlying methods and
the empirical results on real data, see the book 
[Scheuch, Voigt, and Weiss (2023)](https://www.tidy-finance.org), and consult
`?download_data`, `?implement_portfolio_sort`, and `?estimate_fama_macbeth` for
the full set of options at each stage.s

## References
